### 2018-04-01 paper/sexSelectionSpacing-ver1.tex
Edit Distribution of Birth Spacing Across Sex Compositions and Time

I show survival curves conditional on parity progression rather than
standard survival curves. 
The advantage of this approach is that it is possible to directly compare 
the distribution of spacing to next birth across groups, independently of 
differences in how likely the next birth is.
It does, however, depend on covering a long enough spell period
to ensure that only relatively few births are likely to take place after the
censoring date as discussed above.
Because the conditional survival curves are independent of the likelihood
of final parity progression, they all begin at 100\% and end at 0\%.

Instead of averaging across the entire sample, I calculate the conditional
survival curves for an average woman using the method detailed in 
Section \ref{sec:strategy}.
The average woman characteristics are the means of continuous explanatory 
variables and the majority category for categorical explanatory variables.
Common to all is that they do not own land and belong to neither 
a scheduled caste nor a scheduled tribe.

In the interest of brevity, I discuss only a representative subset.
Figure \ref{fig:pps_spell1} shows spacing for the first spell across the
four periods for rural women with no education and urban women with 
eight or more years of education.
Figures \ref{fig:pps_low} and \ref{fig:pps_high} show spacing across sex
compositions for the second, third, and fourth spells for the first period, 
1972--1984, and the last period, 2005--2016, for rural women with no 
education and urban women with eight or more years of education.
The Appendix shows standard and conditional survival curves for all groups and 
time periods, and I discuss some examples of standard survival
curves in the following section.

@@

The duration from marriage to first birth is only affected by son preferences if
people use sex selection to achieve a son as their first-born.
As I discuss above, the biased sex ratio for the first spell for the most educated 
women does suggest that there might be sex selection on the first birth.
I, therefore, begin by comparing in Figure \ref{fig:pps_spell1} how the distribution 
of spacing has changed over time for rural women with no education and urban women 
with eight or more years of education.
What is most striking is how similar the distribution of spacing is across the
two groups.

Both groups show little change in median spacing, but that masks a substantial
compression of when most of the births occur.
In the 1972--1984 period, the middle 80\% of births for women with no education 
are predicted to occur between approximately 6 and 52 months, whereas in the 
2005--2016 period it is between 12 and 46 months.
Hence, the compression is equivalent to a full year.
Women with eight or more years of education show a slightly smaller compression:
in the 1972--1984 period, the middle 80\% of births are predicted to occur between
approximately 6 and 44 months, whereas in the 2005--2016 period it is between
9 and 40 months, or a compression of seven months.

The pattern for both groups shows us that women are less likely to conceive 
before marriage in the later time periods than earlier.
In the first two periods there was a relatively smooth decline in the number of
women without a birth starting at the time of marriage, but in the last two
periods, there are now few women who exit early after marriage and
instead there is a substantial dip between 9 and 12 months after
marriage.

It is likely that the compressed spacing, beginning at nine months
after marriage in the later periods, is associated with better health
and higher age at marriage.
The descriptive statistics above show, for example, that women without 
education have seen an increase in the average age at marriage from below 
16 to 18.5 from the first period to the last while women with the most 
education increased from 19.5 to 20.4.

@@

The distribution of spacing, shown in Figure \ref{fig:pps_high}, for the second spell 
conditional survival curve for urban women with eight or more years of education 
is almost indistinguishable from that of rural women with no education 
for the 1972--1984 period.
The small difference points to these two very different groups of women behaving similarly
in response to son preference in a situation where there was little access to
prenatal sex determination or little incentive to use it early even for those 
who might have had access.

@@

There is an even more pronounced rightward shift for spacing after only
girls in the fourth spell for the first years of the spell. 
The caveat is that there are few births to base the results on,
especially for women with two or more sons.%
\footnote{
As mentioned, it is possible that some women had access to sex selection even
in the first period.
I would expect well educated urban women with more children to be the
most likely to know about and use this early access to prenatal sex determination.
}
Furthermore, the differences are substantial across most of the distribution.
Twelve months after the beginning of the spell there is an approximately ten 
percentage points difference in the conditional survival curves for only girls
and only boys.
That is, of the women who are predicted to have a fourth child just over 30\%
of women with three boys have had their fourth child, while less than 20\% of
those with three girls have had theirs.
Even at three years after the spell, there is a more than five percentage
points difference.

As discussed above there is some evidence of a two-son preference with
the spacing pattern almost the same for women with three boys and with
two boys and one girl.
The conditional survival curve for women with one boy and two girls is
consistently above those with two or more son, but below those with
three girls.
The pattern fit the predicted sex ratios shown above where the highest
predicted sex ratio is for women with only girls, followed by those
with one son and two girls, and those with two or more boys having
almost the same sex ratio.

>>>>

I show survival curves conditional on predicted parity progression rather than
standard survival curves. 
The advantage of this approach is that it is possible to directly compare 
the distribution of spacing to next birth across groups, independently of 
differences in how likely the next birth is.
Because the conditional survival curves are independent of the likelihood
of final parity progression, they all begin at 100\% and end at 0\%.

Instead of averaging across the entire sample, I calculate the conditional
survival curves for an average woman using the method detailed in 
Section \ref{sec:strategy}.
For each combination of education and spell, I use values based on the
average age at the start of the spell and, for the second and higher
spells, the average duration of the first spell.
Furthermore, I use the majority categories for the categorical explanatory 
variables, which means no ownership of land and not in a scheduled caste or 
tribe.
The characteristics used do not change across period to ensure that 
composition effects do not drive the changes.

In the interest of brevity, I discuss only a select set of subsamples.
Figure \ref{fig:pps_spell1} shows spacing for the first spell across the
four periods for rural women with no education and urban women with 
eight or more years of education.
Figures \ref{fig:pps_low} and \ref{fig:pps_high} show spacing across sex
compositions for the second, third, and fourth spells for the first period, 
1972--1984, and the last period, 2005--2016, for rural women with no 
education and for urban women with eight or more years of education.
The Appendix shows standard and conditional survival curves for all groups and 
time periods, and I discuss some examples of standard survival
curves in the following section.%
\footnote{
To ease legibility of the graphs some three months intervals are combined
into longer ones when there are few births available in the sample.
}

@@

The biased sex ratio for the first spell for the most educated 
women suggests that there might be sex selection on the first birth.
I, therefore, begin by comparing in Figure \ref{fig:pps_spell1} how the distribution 
of spacing has changed over time for rural women with no education and urban women 
with eight or more years of education.
What is most striking is how similar the distribution of spacing is across the
two groups.

Both groups show little change in median spacing, but that masks a substantial
compression of when most of the births occur.
In the 1972--1984 period, the middle 80\% of births for women with no education 
are predicted to occur between approximately 6 and 64 months, whereas in the 
2005--2016 period it is between 12 and 54 months.
Hence, the compression is equivalent to more than a full year.
Women with eight or more years of education show a slightly smaller compression:
in the 1972--1984 period, the middle 80\% of births are predicted to occur between
approximately 6 and 48 months, whereas in the 2005--2016 period it is between
12 and 46 months.
Hence, the compression is eight months.

Women became less likely to conceive before marriage over time.
In the first two periods there was a relatively smooth decline in the number of
women without a birth starting at the time of marriage, but in the last two
periods, there are few women who exit early after marriage and
instead there is a substantial dip between 9 and 12 months after
marriage.

It is likely that the compressed spacing, beginning at nine months
after marriage in the later periods, is associated with better health
and higher age at marriage.
For example, women without education have seen an increase in the average 
age at marriage from below 16 to 18.5 from the first period to the last 
while women with the most education increased from 19.5 to 20.4.

@@

The distribution of spacing for the second spell is almost indistinguishable 
between urban women with eight or more years of education and rural women 
with no education for the 1972--1984 period, as shown in Figure \ref{fig:pps_high}.
The small difference points to these two very different groups of women behaving similarly
in response to son preference in a situation where there was little access to
prenatal sex determination or little incentive to use it early even for those 
who might have had access.

@@

There is an even more pronounced rightward shift for spacing after only
girls in the fourth spell for the first years of the spell. 
The caveat is that there are few births to base the results on,
especially for women with two or more sons.%
\footnote{
As mentioned, it is possible that some women had access to sex selection even
in the first period.
I would expect well educated urban women with more children to be the
most likely to know about and use this early access to prenatal sex determination.
}
Furthermore, the differences are substantial across most of the distribution.
Twelve months after the beginning of the spell there is a more than ten 
percentage points difference in the conditional survival curves for only girls
and only boys.
That is, of the women who are predicted to have a fourth child just over 30\%
of women with three boys have had their fourth child, while less than 20\% of
those with three girls have had theirs.
Even at three years after the beginning of the spell, there is a more than 
five percentage points difference.

As discussed above there is some evidence of a two-son preference with
the spacing pattern almost the same for women with three boys and with
two boys and one girl.
The conditional survival curve for women with one boy and two girls is
consistently above those for women with two or more son, but below those 
with three girls.
The pattern fit the predicted sex ratios shown above where the highest
predicted sex ratio is for women with only girls, followed by those
with one son and two girls, and those with two or more boys having
almost the same sex ratio.
